Computing likelihoods under the SMC for a general class of ARGs.

### 2025-02-19 paper.tex

\begin{abstract}
Ancestral recombination graphs (ARGs) are the focus of much ongoing research
interest. Recent progress in inference has made ARG-based approaches feasible
across of range of applications, and many new methods using inferred ARGs as
input have appeared. This progress on the long-standing problem of ARG
inference has proceeded in two distinct directions. The first
class of method in which dramatic progress has been made is in the
statistically rigorous inference of ARGs under the Sequentially Markov
Coalescent (SMC), which are now practical for tens-to-hundreds of samples. 
The second class of recently-developed inference methods are based on
approximate models and heuristics which scale to sample sizes two to three
orders of magnitude larger. Although these heuristic methods are reasonably
accurate under many metrics, one significant drawback is that the ARGs they 
estimate do not have the topological properties required to compute a 
likelihood under models such as the SMC under present-day formulations.
In particular, hueristic inference methods typically do not estimate
precise details about recombination events, which are currently
required to compute a likelihood.
In this paper we present a 
novel backwards-time formulation of the SMC 
and derive a straightforward definition 
of the likelihood of a general class of ARG under this model. 
We show that this formulation does not require precise details of recombination events
to be estimated, and is robust to the presence of polytomies. 
We discuss the possibilities for inference that this opens.
\end{abstract}

@@

\section*{Introduction}
Following recent breakthroughs in simulation and inference methods,
there is now intense interest in applying Ancestral Recombination Graphs (ARGs)
to a variety of questions in evolutionary
biology~\citep{lewanski_era_2024,brandt_promise_2024,nielsen_inference_2025}.
ARGs describe the intricately linked paths
of genetic inheritance resulting from
recombination~\citep{hudson_properties_1983,griffiths_ancestral_1996,wong_general_2023},
and contain rich detail about ancestral processes.
% TODO rephrase this a bit, would be a lot of duplication of references
% in the current formulation.
Numerous applications taking advantage of this detailed history
in inferred ARGs are now appearing~\citep{
stern_approximate_2019,
osmond_estimating_2021,
fan_genealogical_2022,
hejase_deep_2022,
guo_recombination-aware_2022,
ignatieva_ongoing_2022,
wang_complex_2022,
zhang_biobank-scale_2023,
nowbandegani_extremely_2023,
ignatieva_distribution_2023,
fan_likelihood_2023,
huang_estimating_2024,
grundler_geographic_2024,
korfmann_simultaneous_2024,
deraje_inferring_2024,
speidel2025high}
and it seems likely that many more will
follow~\citep{harris_database_2019,harris_using_2023}.
Although there are many different approaches to ARG
inference~\citep{wong_general_2023}, two broad classes of method
have been the focus of most recent interest.

% TODO Maybe swap the order of these two paragraphs around, so it's 
% more natrual to talk about the advantages of model-driven stuff next.
The first class of inference methods are statistically rigorous approaches
that sample ARGs under a population genetics model such as the
coalescent with recombination~\citep{hudson_properties_1983},
and its approximatation, the Sequentially Markovian
Coalescent~\citep{mcvean_approximating_2005,marjoram_fast_2006}.
ARGweaver~\citep{rasmussen_genome-wide_2014} is the most popular,
consistently out-performed other methods in terms 
of accuracy in a variety of benchmarks~\citep{brandt2022evaluation}[MORE
citations,anywhere ARGweaver has been tested for accuracy]
and has succesfully been applied in several
contexts~\citep[e.g.][]{de_chimpanzee_2016,shriner_whole_2018,hejase_genomic_2020,
stankowski_genetic_2024}. 
ARGweaver works [by explanation.]
The recently-introduced SINGER method takes a similar approach [doing something 
slightly different]~\citep{deng_robust_2024}.
While SINGER has not been evaluated as thoroughly as ARGweaver yet, it promises
to be even more accurate while being substantially more
scalable~\citep{deng_robust_2024}.

The second class of inference method that has been of recent interest 
is based on heuristics and approximate models.
Relate~\citep{speidel_method_2019},
tsinfer~\citep{kelleher_inferring_2019},
ARG-Needle~\citep{zhang_biobank-scale_2023} 
and Threads~\citep{gunnarsson_scalable_2024}
work on quite different 
principles, but share some common properties. 
Firstly, they are all 
hueristic and approximation driven, favouring computational
scalability over statistical rigor.
Secondly, they all produce a single
ARG as output, inferring a deterministic point estimate with no means 
of quantifying uncertainty. Thirdly (and most importantly for the 
purposes of this paper) the ARGs they produce lack substantial
detail about ancestral recombination events: the estimates are
rough sketches, hopefully capturing large-scale properties, but lacking
in the finer detail of outputs of ARGweaver or SINGER.

There are substantial advantages to the statistically rigorous sampling 
approaches of ARGweaver and SINGER. Firstly, they tend to be more 
accurate; ARGweaver has consistently out-performed other methods in terms 
of accuracy in a variety of benchmarks~\citep{brandt2022evaluation}[MORE
citations,anywhere ARGweaver has been tested for accuracy]
and, while SINGER has not been evaluated as thoroughly yet, it promises
to be even more accurate~\citep{deng_robust_2024}. 
Secondly, the direct connection with
population genetic models greatly aids interpretability, and allows for
natural extensions like the joint estimation of ARGs and
demography~\citep{hubisz_mapping_2020}. Thirdly, sampling ARGs from a
posterior distribution provides a means of quantifying the uncertainty 
around estimates, which can be used in downstream applications [citation
of this?].

Despite these significant advantages of rigorous inference, however,
hueristic methods remain of interest because of the profound difference
in computational scalability. Although SINGER is a major step forward
in terms of scalability over ARGweaver and other methods, inference is still 
only feasible for hundreds of samples. 
Tsinfer, ARG-Needle and Threads have 
been applied to datasets of hundreds of \emph{thousands} of samples.
With the explosive growth in dataset
sizes in recent years~\citep{caulfield2017national,turnbull2018100,
bycroft2018genome,backman2021exome,halldorsson2022sequences,uk2023whole,
all2024genomic,stark2024call,cook2025our}
there is a pressing need for more statistical
rigor in large scale ARG inference.
% This inability to connect the results of large-scale inference methods
% with a generative model
% is a fundamental limitation, and limits both the utility of 
% current point estimates as well as the possibilities for further
% inference. Specifically, we are unable to quantify uncertainty 
% around heuristic point estimates. We cannot explore ARG space
% around the point estimates using (e.g. MCMC), as we have no way definition
% of improvement. We cannot compare models of demography, or evaluate the 
% goodness of fit for model estimates.
A key problem facing the field is that the ARGs estimated by heuristic large-scale 
methods lack a concrete connection to population genetic theory because
we cannot compute their likelihood under the SMC or similar models. 
% The classical method~\citep{kuhner_maximum_2000} assumes complete knowledge about
% recombination events, and there is no clear way to compute a likelihood for 
% the approximate structures output by hueristic methods.

Computing the likelihood of an ARG under a model is a fundamental element of
probabilistic inference. The classical Kuhner, Yamato \& Felsenstein
\citeyearpar{kuhner_maximum_2000} formulation (henceforth: KYF)
works by assigning a probablility to each recombination and common
ancestor event, and to the inter-event  waiting times. The likelihood of an 
ARG is then a product over all these three classes of term.
% In essence, computing the KYF likelihood requires
% ``replaying'' the  state of Hudson's simulation
% algorithm~\citep{hudson_properties_1983,kelleher_efficient_2016} in order to
% compute the exact configuration of lineages and distribution
% of ancestral material at each time point~\citep{baumdicker_efficient_2021}.
% It therefore requires complete
% detail about each of these events to be stored. 
% Other approaches to defining the likelihood of an ARG under a coalescent-like 
% model follow the same basic event-by-event 
% strategy~\citep[e.g.][]{guo_recombination-aware_2022}.
% % [Are there others?  The bacterial stuff from Didelot etc seem to focus on
% % a tree-by-tree likelihood, rather than global graph, so not clear it's the 
% % same thing]
% Computing the likelihood under the SMC rather than the Hudson coalescent 
% simplifies bookkeeping somewhat,  % Dunno what I'm talking about here
% but a complete event-by-event description of the 
% input ARG is still required.
Although the KYF formulation is a natural and elegant way to describe 
the likelihood of an ARG, such complete detail is not present 
in many estimated ARGs~\citep{wong_general_2023} or even in most simulations.
Simulators such as the classical \texttt{ms}
program~\citep{hudson_generating_2002} output ARGs in a tree-by-tree format,
which does not contain sufficient detail for likelihood computation.
The \texttt{msprime}
simulator~\citep{kelleher_efficient_2016,baumdicker_efficient_2021} 
by default outputs a more complete and efficient representation of 
the simulated ancestry, but still does not provide the exhaustive
detail required.
Indeed, providing the information required to 
support likelihood calculations originally motivated the addition of
the \texttt{record\_full\_arg}
option to \texttt{msprime}~\citep{baumdicker_efficient_2021}.
Forwards-time simulations that output 
ARGs~\citep{kelleher_efficient_2018,haller_tree_2018} similarly do not
typically contain the level of detail required to compute a likelihood under
the KYF approach.

@@

In this paper we provide a new formulation of ARG likelihood which 
has much more lenient input requirements than KYF, and gracefully handles
incomplete and imprecise ARGs.
% In this paper we provide a necessary first step towards this goal
% of statistically rigorous large-scale ARG inference, by defining a function
% to compute the likelihood of a general class of ARG under the SMC.
To do this, we rederive the SMC under a backwards-time formulation,
modelling the process in terms of lineages and [something] rather
than events. Using this novel formulation, we then show how to 
easily compute the likelihood of any ARG under the SMC and outline 
an efficient algorithm to do so. We illustrate the flexibility of this
approach via [some examples] and conclude by discussing avenues
for future research opened up.

@@

\begin{figure}
\centering
\includegraphics[width=0.9\textwidth]{figures/smc_custom_2rows_area_full_hap.png}
\caption{ARG generated under the SMC, represented as a graph (top) and as a
series of local trees (bottom). To ensure the one-to-one correspondence
between both representations a node is recorded for each local tree whenever
a node is encountered along its ancestry path in the graph representation.
This results in nodes that (locally) have only a single child, and are therefore
(locally) unary. 
Nodes A, B and C are recombinants, i.e, have two parents in the graph
and are unary in the local trees. The blue shaded regions indicate
the range of possible times associated with these recombinants.
The blue arrows show the
standard left-to-right logic of the SMC whereby
the floating lineage coalesces randomly
with a the remaining portion of the local tree following recombination.
% By removing recombinant nodes, we limit the information on the time to each
% of these events to the windows (blue shaded array) delimited by the age of
% the child node and the minimum age of its parents.
}
\label{fig:smc-unary}
\end{figure}

@@

Linking the resulting tree-valued process along the genome back to
the relationship between the ``graph'' and
``local trees'' representation of ARGs, 
the Markovian property can be formulated as follows.
The distribution of the next tree within the series
of local trees can be generated only knowing the current tree,
provided that we remove recombinant nodes entirely,
and erase the unary portions of coalescent nodes.
To see that the locally unary nodes make the process non-Markov,
consider that a node in a tree might be unary either because
it's about to be (in the left-to-right sense) a coalescent node 
or it already has been:
looking at the second tree in Fig.\ref{fig:smc-unary},
we know node $f$ will be coalescent further along the sequence
because it was not present in the first tree.
So, the first tree gives us information about subsequent trees
that the second tree alone could not.
Further note that by limiting common ancestry events to overlapping 
pairs of lineages, we are guaranteed that all nodes in the ARG 
only carry ancestral material, that is,
genetic material that is inherited by either one of a set of samples
\citep{wiuf_ancestry_1999}.
We therefore require a single half-open interval to describe
the ancestral
material associated with each node in the ARG.

@@

Assuming a randomly mating, diploid population of constant size $N_e$,
common ancestry occurs between
two \emph{overlapping} lineages at rate $\lambda = 1/(2N_e)$.
Here we deviate from the McVean and Cardin formulation,
and define a strict total order on all lineages in $L(t)$:
for each lineage $X_j(t) = [x_{j}, y_{j})$ we define its 
order by the left coordinate $x_j$ and label $j$ (to break ties).
The instantaneous rate of coalescence then equals $\lambda$ 
multiplied by the number of overlapping pairs,
i.e., $\lambda \sum_{i \neq j} I_{jk}$,
where
\begin{equation} \label{def:coal}
I_{jk} = \begin{cases}
    1 & X_j > X_k \text{ and } y_k > x_j \\
    0 & \text{otherwise.}
\end{cases}
\end{equation}
We say that lineage $k$ can coalesce \emph{into} lineage $j$ if $I_{jk} = 1$.
From the definition of the strict total order it follows that if $I_{jk} = 1$
then $I_{kj} = 0$. 
The newly formed lineage acquires the union of both intervals,
the original lineages are removed, and $L(t)$ is updated accordingly.
% [JK notes: need to explain here that this deliberately introduced 
% asymmetry doesn't affect the actual process. Also need to comment on 
% how the process terminates. Basically has the same subtleties as McVean and 
% Cardin, and not important for our purposes.]
For later use, also define $C_j(t)$ to be the set of
lineages that can coalesce into lineage $j$, i.e., 
$C_j(t) = \{X_k \in L(t) | I_{jk}(t) = 1\}$; 
thus, $|C_j(t)| = I_{j}(t) = \sum_{k} I_{jk}(t)$,
and the total rate of coalescence at time $t$ is $\sum_{j} |C_j(t)|$. 
Because $I_{jk}$ is defined in terms of a total order, the sets $C_j(t)$ are disjoint,
which will simplify likelihood computations later. In particular, although
recombination affects \emph{which} lineages can coalesce into $j$, it does not
change the \emph{number} of such lineages: in other words, a recombination
event changes $C_j(t)$ but not $I_{j}(t)$.

@@

% Finally, by describing the SMC
% backwards in time, some intervals contained within the half-open
% interval representing a lineage might have fully coalesced before all the genetic
% material carried by that lineage has. As noted by \citet{mcvean_approximating_2005},
% such intervals can be observed as the stalks on the marginal genealogies (e.g.\ marginal trees
% for $[25, 81)$ Fig. \ref{fig:smc-unary}). Recombination events occurring within these smaller
% intervals however do not affect the marginal genealogies.

@@

\begin{figure}
    \centering
    \includegraphics[width=0.65\textwidth]{figures/single_event.png}
    \caption{
    Graphical representation of the events associated with the edge $(p,c,x,y)$.
    The lineage carried by $c$ (dark green bar) spans $[x_c,y)$ 
    until a recombination event (blue
    dotted line) splits it into two segments. Left and right hand segments coalesce
    (with the pink and yellow lineages, respectively)
    at times $t_{p'}$ and $t_p$ respectively, bounding the time to recombination by
    % JK used to be: $(c. \min(t_p, t_{p'})$
    $\min(t_p, t_{p'}$.
    % JK: not sure this helps? 
    % Note that the contribution to the recombination hazard
    % by edge $(p,c,x,y)$ only covers ``half'' of the green shaded area. The other
    % half is contributed by edge $(p', c, x_c, x)$. 
    \label{fig:likelihood}
    }
\end{figure}

Now consider the ``hazard'' associated with recombination.
Begin with an edge $e = (p,c,x,y)$.
This implies that there was a sequence of lineages
from the time of the child $c$ (call this time $t_c$)
back to the time of the parent $p$ (again, $t_p$),
that contained the segment $[x,y)$,
and the lineage that directly inherited from $p$
had the segment $[x,y)$ (see Fig.~\ref{fig:likelihood}).
We therefore know there was no recombination between $x$ and $y$
on any of those lineages 
over this time span,
the probability of which is
$\exp(-r \times \mathcal{A})$, where $r$ is the recombination rate
and $\mathcal{A}$ is the total area (eligible links times depth) of the edge.
So, the first contribution to the likelihood from this edge $e$ is
\begin{equation}\label{eq:span}
A_e(r) = \exp(-r (y-x - \mathbbm{1}_{x > x_c})(t_p - t_{c})) .
\end{equation}
Note that if a recombination did occur, then the number of links not
affected by recombination is 1 less (see eq.~\eqref{eq:depth}).

@@

Recall that we define a total order on edges,

@@

and the smaller of $t_p$ and $t_{p'}$. Note that two events must have happened
between $t_c$ and $\min(t_p, t_{p'})$; taking all this into consideration,

@@

        & \qquad \text{if } x>x_{c} ,

@@

where $x_c$ is the leftmost position of the edges whose children are $c$.

@@

The overall likelihood of an ARG defined by the set of edges $E$ and nodes $N$

@@

    (r)^{|E|-|N|+n_r} (\lambda)^{|E|-|N|+n_s} \prod_{(p, c) \in E} A_e(r) B_e(r, \lambda) .

@@

The likelihood of a set of mutations given an ARG is

@@

    \frac{1}{\mathcal{M}!} \prod_{(p, c, x, y) \in E} e^{-(y-x)(t_p-t_c)\mu^{|m_e|}}

@@

where $\mathcal{M}$ the total number of mutations and $m_e$
the number of mutations along edge $(c,p)$ \citep{mahmoudi_bayesian_2022}.

@@

Eq.~\eqref{eq:full-lik} allows us to compute a likelihood for \emph{any}
collection of nodes (ancestral haplotypes that lived at some time 
$t$ in the past) and edges (a record that child node $c$ 
inherited the genomic interval $[x, y)$ from parent node $p$)
under the SMC. Other than some basic consistency constraints (e.g.\ parents
must be older than children) there are no requirements on the topology
described, and in particular, no requirement that the events from the 
model (the SMC) should be explicitly identified in the graph. Our 
formulation reasons about the events that \emph{could have resulted in}
the observed nodes and edges, and therefore frees us from the requirement
of having to explicitly write those down. This gives a great deal 
of flexibility, which we illustrate in this section via some examples.

>>>>

\begin{abstract}
Ancestral recombination graphs (ARGs) are the focus of much ongoing research
interest. Recent progress in inference has made ARG-based approaches feasible
across of range of applications, and many new methods using inferred ARGs as
input have appeared. This progress on the long-standing problem of ARG
inference has proceeded in two distinct directions. First,
the Bayesian inference of ARGs under the Sequentially Markov
Coalescent (SMC), is now practical for tens-to-hundreds of samples. 
Second, approximate models and heuristics can now scale to sample sizes two to three
orders of magnitude larger. Although these heuristic methods are reasonably
accurate under many metrics, one significant drawback is that the ARGs they 
estimate do not have the topological properties required to compute a 
likelihood under models such as the SMC under present-day formulations.
In particular, hueristic inference methods typically do not estimate
precise details about recombination events, which are currently
required to compute a likelihood.
In this paper we present a 
backwards-time formulation of the SMC 
and derive a straightforward definition 
of the likelihood of a general class of ARG under this model. 
We show that this formulation does not require precise details of recombination events
to be estimated, and is robust to the presence of polytomies. 
We discuss the possibilities for inference that this opens.
\end{abstract}

@@

\section*{Introduction}
Following recent breakthroughs in simulation and inference methods,
there is now strong interest in applying Ancestral Recombination Graphs (ARGs)
to a variety of questions in evolutionary
biology~\citep{lewanski_era_2024,brandt_promise_2024,nielsen_inference_2025}.
ARGs describe the intricately linked paths
of genetic inheritance resulting from
recombination~\citep{hudson_properties_1983,griffiths_ancestral_1996,wong_general_2023},
and contain rich detail about ancestral processes.
% TODO rephrase this a bit, would be a lot of duplication of references
% in the current formulation.
Numerous applications taking advantage of this detailed history
in inferred ARGs are now appearing~\citep{
stern_approximate_2019,
osmond_estimating_2021,
fan_genealogical_2022,
hejase_deep_2022,
guo_recombination-aware_2022,
ignatieva_ongoing_2022,
wang_complex_2022,
zhang_biobank-scale_2023,
nowbandegani_extremely_2023,
ignatieva_distribution_2023,
fan_likelihood_2023,
huang_estimating_2024,
grundler_geographic_2024,
korfmann_simultaneous_2024,
deraje_inferring_2024,
speidel2025high}
and it seems likely that many more will
follow~\citep{harris_database_2019,harris_using_2023}.
Although there are many different approaches to ARG
inference~\citep{wong_general_2023}, two broad classes of method
have been the focus of most recent interest.

% TODO Maybe swap the order of these two paragraphs around, so it's 
% more natrual to talk about the advantages of model-driven stuff next.
The first class of inference methods are Bayesian approaches
that sample ARGs under a population genetics model such as the
coalescent with recombination~\citep{hudson_properties_1983},
and its approximatation, the Sequentially Markovian
Coalescent~\citep{mcvean_approximating_2005,marjoram_fast_2006}.
ARGweaver~\citep{rasmussen_genome-wide_2014} has been the most widely used and studied,
and until recently consistently out-performed other methods in terms 
of accuracy in a variety of benchmarks~\citep{brandt2022evaluation},
and has succesfully been applied in several
contexts~\citep[e.g.,][]{de_chimpanzee_2016,shriner_whole_2018,hejase_genomic_2020,
stankowski_genetic_2024}. 
The recently-introduced method SINGER takes a similar approach,
and by using an improved Monte Carlo algorithm
promises to be even more accurate while being substantially more
scalable~\citep{deng_robust_2024}.

The second class of inference method that has been of recent interest 
is based on heuristics and approximate models.
Relate~\citep{speidel_method_2019},
tsinfer~\citep{kelleher_inferring_2019},
ARG-Needle~\citep{zhang_biobank-scale_2023} 
and Threads~\citep{gunnarsson_scalable_2024}
work on quite different 
principles, but share some common properties. 
Firstly, they are all 
hueristic and approximation driven, favouring computational
scalability over a basis in a statistical model.
Secondly, they all produce a single
ARG as output, inferring a deterministic point estimate with no means 
of quantifying uncertainty. Thirdly (and most importantly for the 
purposes of this paper) the ARGs they produce 
lack some of the information present in ARGweaver or SINGER. 
This occurs, roughly speaking, because of a fundamental difference in approach:
not all details of the ARG can be reliably inferred
(e.g., the timing of recombination events, as discussed below);
and the heuristic methods deal with this uncertainty by omitting these events,
while ARGweaver and SINGER provide a sample from the posterior.
(So, some aspects of the resulting ARGs should be interpreted slightly differently;
see \citet{wong_general_2023} for more discussion.)

There are substantial advantages to the statistically rigorous sampling 
approaches of ARGweaver and SINGER.
Firstly, they tend to be more accurate when compared on the same data set.
Secondly, the direct connection with population genetic models allows for
natural extensions like the joint estimation of ARGs and
demography~\citep{hubisz_mapping_2020}.
Thirdly, sampling ARGs from a
posterior distribution provides a means of quantifying the uncertainty 
around estimates, which can be used in downstream applications.

On the other hand, heuristic methods have the potential
to utilize the greater amount of information in large datasets,
thus possibly producing even more accurate estimates
(although this has not, to our knowledge, been tested).
Although SINGER is a major step forward
in terms of scalability over ARGweaver and other methods, inference is still 
only feasible for hundreds of samples. 
Meanwhile, tsinfer, ARG-Needle and Threads have 
been applied to datasets of hundreds of \emph{thousands} of samples.
With the explosive growth in dataset
sizes in recent years~\citep{caulfield2017national,turnbull2018100,
bycroft2018genome,backman2021exome,halldorsson2022sequences,uk2023whole,
all2024genomic,stark2024call,cook2025our}
there is a pressing need for more statistical
rigor in large scale ARG inference.
% This inability to connect the results of large-scale inference methods
% with a generative model
% is a fundamental limitation, and limits both the utility of 
% current point estimates as well as the possibilities for further
% inference. Specifically, we are unable to quantify uncertainty 
% around heuristic point estimates. We cannot explore ARG space
% around the point estimates using (e.g., MCMC), as we have no way definition
% of improvement. We cannot compare models of demography, or evaluate the 
% goodness of fit for model estimates.
A key problem facing the field is that the ARGs estimated by heuristic large-scale 
methods lack a concrete connection to population genetic theory because
we cannot compute their likelihood under the SMC or similar models. 
% The classical method~\citep{kuhner_maximum_2000} assumes complete knowledge about
% recombination events, and there is no clear way to compute a likelihood for 
% the approximate structures output by hueristic methods.

Computing the likelihood of an ARG under a model is a fundamental element of
probabilistic inference. The classical Kuhner, Yamato \& Felsenstein
\citeyearpar{kuhner_maximum_2000} formulation (henceforth: KYF)
works by assigning a probablility to each recombination and common
ancestor event, and to the inter-event  waiting times. The likelihood of an 
ARG is then a product over all these three classes of term.
% In essence, computing the KYF likelihood requires
% ``replaying'' the  state of Hudson's simulation
% algorithm~\citep{hudson_properties_1983,kelleher_efficient_2016} in order to
% compute the exact configuration of lineages and distribution
% of ancestral material at each time point~\citep{baumdicker_efficient_2021}.
% It therefore requires complete
% detail about each of these events to be stored. 
% Other approaches to defining the likelihood of an ARG under a coalescent-like 
% model follow the same basic event-by-event 
% strategy~\citep[e.g.,][]{guo_recombination-aware_2022}.
% % [Are there others?  The bacterial stuff from Didelot etc seem to focus on
% % a tree-by-tree likelihood, rather than global graph, so not clear it's the 
% % same thing]
% Computing the likelihood under the SMC rather than the Hudson coalescent 
% simplifies bookkeeping somewhat,  % Dunno what I'm talking about here
% but a complete event-by-event description of the 
% input ARG is still required.
Although the KYF formulation is a natural and elegant way to describe 
the likelihood of an ARG, such complete detail is not present 
in many estimated ARGs~\citep{wong_general_2023} or even in most simulations.
Simulators such as the classical \texttt{ms}
program~\citep{hudson_generating_2002} output ARGs in a tree-by-tree format,
which does not contain sufficient detail for likelihood computation.
The \texttt{msprime}
simulator~\citep{kelleher_efficient_2016,baumdicker_efficient_2021} 
by default outputs a more complete and efficient representation of 
the simulated ancestry, but still does not provide the exhaustive
detail required.
Indeed, providing the information required to 
support likelihood calculations originally motivated the addition of
the \texttt{record\_full\_arg}
option to \texttt{msprime}~\citep{baumdicker_efficient_2021}.
Forwards-time simulations that output 
ARGs~\citep{kelleher_efficient_2018,haller_tree_2018} similarly do not
typically contain the level of detail required to compute a likelihood under
the KYF approach.

@@

In this paper we provide a new formulation of ARG likelihood which 
has much more lenient input requirements than KYF, and gracefully handles
incomplete and imprecise ARGs.
% In this paper we provide a necessary first step towards this goal
% of statistically rigorous large-scale ARG inference, by defining a function
% to compute the likelihood of a general class of ARG under the SMC.
To do this, we give a backwards-in-time description of the SMC
which naturally leads to a straightforward method for computing the likelihood
of any ARG, and describe an efficient algorithm to do so.
We also illustrate the utility of this approach via some simple examples.

@@

\begin{figure}
\centering
\includegraphics[width=0.9\textwidth]{figures/smc_custom_2rows_area_full_hap.png}
\caption{An ARG generated under the SMC, represented as a graph (top) and as a
series of local trees (bottom). To ensure the one-to-one correspondence
between both representations a node is recorded for each local tree whenever
a node is encountered along its ancestry path in the graph representation.
This results in nodes that (locally) have only a single child, and are therefore
(locally) unary. 
Nodes A, B and C are recombinants, i.e, have two parents in the graph
and are unary in the local trees. The blue shaded regions indicate
the range of possible times associated with these recombinants.
The blue arrows show the
standard left-to-right logic of the SMC whereby
the floating lineage coalesces randomly
with a the remaining portion of the local tree following recombination.
% By removing recombinant nodes, we limit the information on the time to each
% of these events to the windows (blue shaded array) delimited by the age of
% the child node and the minimum age of its parents.
}
\label{fig:smc-unary}
\end{figure}

@@

Linking the resulting tree-valued process along the genome back to
the relationship between the ``graph'' and
``local trees'' representation of ARGs, 
the Markovian property can be formulated as follows.
The distribution of the next tree within the series
of local trees can be generated only knowing the current tree,
provided that we remove recombinant nodes entirely,
and erase the unary portions of coalescent nodes.
To see that the locally unary nodes make the process non-Markov,
consider that a node in a tree might be unary either because
it's \emph{about to be} (in the left-to-right sense) a coalescent node 
or it \emph{already has been}:
looking at the second tree in Fig.\ref{fig:smc-unary},
we know node $f$ will be coalescent further along the sequence
because it was not present in the first tree.
So, the first tree gives us information about subsequent trees
that the second tree alone could not.
Further note that by limiting common ancestry events to overlapping 
pairs of lineages, we are guaranteed that all nodes in the ARG 
only carry ancestral material, that is,
genetic material that is inherited by either one of a set of samples
\citep{wiuf_ancestry_1999}.
We therefore require a single half-open interval to describe
the ancestral
material associated with each node in the ARG.

@@

Assuming a randomly mating, diploid population of constant size $N_e$,
common ancestry occurs between
two \emph{overlapping} lineages at rate $\lambda = 1/(2N_e)$.
Here we deviate from the McVean and Cardin formulation,
and define a strict total order on all lineages in $L(t)$:
for each lineage $X_j(t) = [x_{j}, y_{j})$ we define its 
order by the left coordinate $x_j$ and label $j$ (to break ties).
The instantaneous rate of coalescence then equals $\lambda$ 
multiplied by the number of overlapping pairs,
i.e., $\lambda \sum_{i \neq j} I_{jk}$,
where
\begin{equation} \label{def:coal}
I_{jk} = \begin{cases}
    1 & X_j > X_k \text{ and } x_j < y_k \\
    0 & \text{otherwise.}
\end{cases}
\end{equation}
We say that lineage $k$ can coalesce \emph{into} lineage $j$ if $I_{jk} = 1$.
Note that since $X_j > X_k$ implies that $x_j \ge x_k$,
this condition implies that the left endpoint $x_j$ of $X_j$
falls in the interval defined by $X_k$.
From the definition of the strict total order it follows that if $I_{jk} = 1$
then $I_{kj} = 0$. 
The newly formed lineage acquires the union of both intervals,
the original lineages are removed, and $L(t)$ is updated accordingly.
% [JK notes: need to explain here that this deliberately introduced 
% asymmetry doesn't affect the actual process. Also need to comment on 
% how the process terminates. Basically has the same subtleties as McVean and 
% Cardin, and not important for our purposes.]
For later use, also define $C_j(t)$ to be the set of
lineages that can coalesce into lineage $j$, i.e., 
$C_j(t) = \{X_k \in L(t) | I_{jk}(t) = 1\}$; 
thus, $|C_j(t)| = I_{j}(t) = \sum_{k} I_{jk}(t)$,
and the total rate of coalescence at time $t$ is $\sum_{j} |C_j(t)|$. 
Because $I_{jk}$ is defined in terms of a total order, the sets $C_j(t)$ are disjoint,
which will simplify likelihood computations later. In particular, although
recombination affects \emph{which} lineages can coalesce into $j$, it does not
change the \emph{number} of such lineages: in other words, a recombination
event changes $C_j(t)$ but not $I_{j}(t)$.

@@

% Finally, by describing the SMC
% backwards in time, some intervals contained within the half-open
% interval representing a lineage might have fully coalesced before all the genetic
% material carried by that lineage has. As noted by \citet{mcvean_approximating_2005},
% such intervals can be observed as the stalks on the marginal genealogies (e.g., marginal trees
% for $[25, 81)$ Fig. \ref{fig:smc-unary}). Recombination events occurring within these smaller
% intervals however do not affect the marginal genealogies.

@@

\begin{figure}
    \centering
    \includegraphics[width=0.65\textwidth]{figures/single_event.png}
    \caption{
    Graphical representation of the events associated with the edge $(p,c,x,y)$.
    The lineage carried by $c$ (dark green bar) spans $[x_c,y)$ 
    until a recombination event (blue
    dotted line) splits it into two segments. Left and right hand segments coalesce
    (with the pink and yellow lineages, respectively)
    at times $t_{p'}$ and $t_p$ respectively, bounding the time to recombination by
    % JK used to be: $(c. \min(t_p, t_{p'})$
    $\min(t_p, t_{p'})$.
    % JK: not sure this helps? 
    % Note that the contribution to the recombination hazard
    % by edge $(p,c,x,y)$ only covers ``half'' of the green shaded area. The other
    % half is contributed by edge $(p', c, x_c, x)$. 
    \label{fig:likelihood}
    }
\end{figure}

Now consider the ``hazard'' associated with recombination.
Begin with an edge $e = (p,c,x,y)$.
This implies that there was a sequence of lineages
from the time of the child $c$ (call this time $t_c$)
back to the time of the parent $p$ (again, $t_p$),
that contained the segment $[x,y)$,
and the lineage that directly inherited from $p$
had the segment $[x,y)$ (see Fig.~\ref{fig:likelihood}).
We therefore know there was no recombination between $x$ and $y$
on any of those lineages 
over this time span,
the probability of which is
$\exp(-r \mathcal{A}_e)$, where $r$ is the recombination rate
and $\mathcal{A}_e$ is the total area (eligible links multiplied by length in time) of the edge.
The length in time of the edge is just $t_p - t_c$,
while the number of ``eleigible links'' in the edge is the number of adjacent base pairs in the edge
on which recombination did not occur, which is either $y-x$
(if this edge is the leftmost of all edges above node $c$)
or $y-x-1$ (otherwise).
So, the first contribution to the likelihood from this edge $e$ is
\begin{equation}\label{eq:span}
A_e(r) = \exp(-r (y-x - \mathbbm{1}_{x > x_c})(t_p - t_{c})) ,
\end{equation}
where as for lineages, $x_c$ is the leftmost position of the edges whose children are $c$.
% Note that if a recombination did occur, then the number of links not
% affected by recombination is 1 less (see eq.~\eqref{eq:depth}).

@@

Following what we did for lineages, we define a total order on edges

@@

and the smaller of $t_p$ and $t_{p'}$.
Taking all this into consideration,

@@

        & \qquad \text{if } x>x_{c} .

@@

@@

The overall likelihood of an ARG $\mathcal{G}$ defined by the set of edges $E$ and nodes $N$

@@

    (r)^{|E|-|N|+n_r} (\lambda)^{|E|-|N|+n_s} \prod_{e \in E} A_e(r) B_e(r, \lambda) .

@@

The likelihood of a set of mutations given an ARG $\mathcal{G}$ given mutation rate $\mu$ is

@@

    \prod_{e \in E} \frac{(\mu(t_p-t_c)(y-x))^{m_e}}{m_e!} e^{-\mu (t_p-t_c)(y-x)} ,
    % I don't think the following is correct:
    % \frac{1}{\mathcal{M}!} \prod_{(p, c, x, y) \in E} e^{-(y-x)(t_p-t_c)\mu^{|m_e|}}

@@

where 
% $\mathcal{M}$ the total number of mutations and 
$m_e$ is the number of mutations along edge $e$ 
and as before $e = (p,c,x,y)$
\citep{mahmoudi_bayesian_2022}.

@@

Eq.~\eqref{eq:full-lik} allows us to compute a likelihood for \emph{any}
collection of nodes (ancestral haplotypes that lived at some time 
$t$ in the past) and edges (a record that child node $c$ 
inherited the genomic interval $[x, y)$ from parent node $p$)
under the SMC. Other than some basic consistency constraints (e.g., parents
must be older than children) there are no requirements on the topology
described, and in particular, no requirement that the events from the 
model (the SMC) should be explicitly identified in the graph. Our 
formulation reasons about the events that \emph{could have resulted in}
the observed nodes and edges, and therefore frees us from the requirement
of having to explicitly write those down. This gives a great deal 
of flexibility, which we illustrate in this section via some examples.
